Search for Spatially Extended Fermi-LAT Sources Using Two Years of Flight Data

### 2011-06-29 extended_catalog/paper/extended_catalog.tex
typos and stuff

@@

\begin{abstract}
We present a new method for analyzing spatially extended sources with
the Fermi Large Area Telescope (LAT), the primary science instrument
on the {\em Fermi Gamma-ray Space Telescope (Fermi)}.  We provide a
series of Monte Carlo studies that were done to validate this tool.
We then calculate the Fermi LAT's sensitivity to spatially exnteded
sources.  We then apply this tool to test all sources in the 2FGL Two
Year Catalog for extension. We then report on the discovery
of XXXXXX\todo{What number should be put
here???} new extended sources.
\end{abstract}

@@

The Large Area Telescope (LAT) is a pair conversion telescope on the
Fermi Gamma-Ray Space Telescope. It has been surveying the $\gamma$-ray
sky since June 2008.  Fermi is particularly well suited as a survey
instrument because of its to survey the broad energy coverage (20 \mev
to >300 \gev), a wide field of view ($\sim 2.4 \sr$) and large effective
area ($\sim 8000 \cm^2$ at $>1 \gev$).

Using one year of all sky monitoring, the LAT published a catalog of
source that were significantly resolved.  Many of these sources could
be be associated with a variety of source classes including active
galactic nuclei, pulsars, galaxies, and supernova remenants.  Many of
these source classes can be spatially resolved when observed at other
frequencies and could possibly be spatially resolvable by the LAT.

In particular, the LAT has discovered a variety of spatially extended
supernova remnants including IC443 and W51C (\cite{ic443,w51c}) and pulsar
wind nebulae (\cite{msh1552,velax}). And there are other source classes
which could possibly be spatially resolved by the LAT including galaxy
clusters, dark matter, and possibly new physics.

Furthermore, a variety of spatially extended but otherwise unassociated
sources have been detected in the galacitc plane at \tev energies using
Air Cherenkov detectors. Most prominent was a survey of the galaictc
plane using H.E.S.S which discovered 14 spatially extended sources
(\cite{HESS_plane_survey}).

Being able to spatially resolve a fermi source is important for several
reasons. Most importantly, the spatial morphology can be used to help
identify a Fermi source. Because of the large point spread function (PSF),
for any source there are often several counterparts which it could be
associated with. Finding a matching morphology can uniquly identify the
source emitting the $\gamma$-rays.

Because of the largely varying PSF of the LAT, the spatial and spectral
model of a source do not nicely decouple and spatially resolving a Fermi
source is also important for obtaining an unbiased spectral model of
the source.  Furthermore, using this extension information is important
for improved the overall model of the sky which is important for obtaining
a better fit of nearby sources and removing spurious point sources which
we might otherwise think we detect.

Nevertheless, morphological studies of sources in the \gev energy range
using the Fermi LAT is challenging because of the significantly energy
dependent PSF and because of systematic errors in the galactic diffuse
emission.  The Fermi PSF is determinted at lower energies by multiple
scatteering in the tracker The 68\% containment radius of the PSF is
approximatly 5.1\deg at 100 \mev  when averaged over instrumetn acceptance
and including photons which convert in either the thick or thin layers of
the tracker.  The PSF approaches a limit given by the granularity of the
tracker channeles and is 0.14\deg at 10 \gev (\cite{on_orbit_calibration}).
Therefore, the higher energy photons are significantly more important for
resolving a source but there are typically fewere of them.  So a detailed
maximum likelihood analysis must be used to maximize our sensitivity.

@@

@@

A new analysis tool has been developed to address the unique requirements
for studying spatially extended sources with the Fermi LAT. The tool
implements a maximum likelihood analysis where the Poisson likelihood
for observing the measured counts is computed for a particular sky
model. The parameters of the sky model are then fit by maximizing the log
of the likelihood.  To study the extension of a source, one can assume
a particular shape for an extended source (e.g. a disk or Gaussian),
and fit the spatial parameters of that source.

In principle, the entire task can be done using the official maximum
likelihood tool \gtlike which is distributed publically by the Fermi
Science Support Center\cite{fssc}.  But \gtlike is only able to fit the
spectral parameters of the source, so the extension fitting would have
to be done as an external loop. Typically, the run time for a binned
source analysis using \gtlike is several hours so this loop would be
especially time consuming. In principle, parallel computing could ease
this burdon, but parallel fitting algorithms are rather untested and
complex. Furthermore, the machinery required to do something like this
would necessarily be rather complex. What has typically been done by
the Fermi team to study extended sources is to fix the assumed position
of the extended source, either by our knowledge of a source from other
wavelenghts, or by fitting it as a point source. Then, a profile of the
likelihood as a function of extension is developed using \gtlike and the
maximum of this profile. This approach is not optimal because it needs
not be the case that the best fit position of a point source would be
the same as the best fit position of an extended source. Furthermore, by
not fully maximizing the likelihood, this method is not as significant to
extension, especially of faint sources. And finally, since this approach
is so computationally intensive, no large scale Monte Carlo effort has
been launched to validate the method.

Instead, an alternate approach was developed for performing a
morphological analysis. An alternate maximum likelihood fitting package
called \pointlike has been developed by the Fermi collaboration and we
have expanded its functionality to fully support analyzing spatially
extended sources. \pointlike is most thoroughly described in Matthew
Kerr's Ph.D Thesis but the important aspects will be summarized below
(\cite{matthew_kerr_thesis}) . What differentiates \pointlike most
significantly from \gtlike is its emphasis on speed by introducing
approximations into the calculation of the likelihood function while
at the same time trying to reduce any numerical errors that may
be introduced.  Like binned \gtlike, \pointlike bins the sky both in
position and energy.  Unlike \gtlike, \pointlike relies on the {\em
healpix} representation of spatial bins\cite{healpix} and scales the
size of each spatial bin to be smaller than the PSF. This is convenient
because it no significant information is lost due to binning at all
energies and it is always a good approximation that the model predicted
counts are uniform across a spatial bin.

\pointlike additionaly bins the Fermi $\gamma$-rays by their conversion
type, which corresponds to whether the photon converted in the front or
back part of the detector. Since the PSF is smaller for front entering
photons, one would expect binning each conversion type independently to
improve our sensitivity.

Furthermore, \pointlike uses a sparse matrix representation of the spatial
bins where only the bins with counts in them are looped over. The only
downside to this is that the integral of the model predicted counts
must be independently calculated.  Effectively, \pointlike efficiently
interpolates from a totally binned analysis at low energy (where the
resolution is poor and there are many counts) to an unbinned analysis
at high energy (where the resolution is very good but there are few
counts). This approach provides a significant savings in time while
introducing only small numerical error.

Finally, \pointlike is written mostly in the \python programming language
using numpy and scipy to vectorize most of the computations. This provides
a reasonable tradeoff between computational efficiency and quickness of
development time. And by leaving the interface to use \pointlike as a
series of \python modules, advanced functions including extension fittion
described in section~\ref{extesion_fitting} and the dual localization
procedure described in section~\ref{dual_localization_method} could
quickly be build on top of the core modules.

@@

We have modified \pointlike to support fitting spatially extended sources.
To do this, we introduced a new class of diffuse sources where the spatial
and spectral information completely decouple and where the spatial shape
is parameters by simple shapes.  The code consistently convolves the
extended source shape with the point spread function (as a function
of energy) and integrates the spectrum over energy to computed model
predicted counts.

Finally, \pointlike allows the spatial parameters of an extended source
to be simulataneoulsy fit. \pointlike uses the \minuit fitting library to
perform the maximum likelihood maximization of the position and extension
of an extended source (\cite{minuit_documentation}).  At each step in
the fit of the shape, the spectral parameters of the entire sky model
are reoptimized. To avoid projection effects, what is directly fit is
not the longitude and latitude of the source but instead the displacement
in a rotated reference frame.

\pointlike optimizes the analysis of extended sources in two ways.  First,
for extended sources which are radially symmetric, the convolution of
the source shape with our parameterization of the PSF can be performed
semi-analytically. Generally, at a given energy the observed photon
distribution is
\begin{equation}
  \text{PDF}(\vec r) = \int  \text{PSF}(|\vec r - \vec r'|)I_\text{src}(\vec r') d A' 
\end{equation}
where $I_\text{src}$ is the extended soruce shape and the PSF is parameterized by
\begin{equation}
  \text{PSF}(\vec r) = 
  \frac{1}{2\pi\sigma^2}
  \left(1-\frac{1}{\gamma}\right)
  \left(1+\frac{u}{\gamma}\right)^{-\gamma}
\end{equation}
Here, $\sigma$ and $\gamma$ are parameterize the $PSF$.  If the extended
source is radially symmetric, then $I_\text{src} (\vec r') = I_\text{src}
(r')$ and the angular part of the integral can be done analytically,
leaving a simpler one dimensional integral:
\begin{equation}
  \text{PDF}(u)= \int_0^\infty dv
  I_\text{src}(v) 
  \left(\frac{\gamma-1}{\gamma}\right)
  \left( \frac{\gamma}{\gamma + u + v}\right)^\gamma 
  \times ~_2F_1 \left(\gamma/2,\frac{1+\gamma}{2},1,\frac{4uv}{(\gamma+u+v)^2}\right)
\end{equation}
Not only is this convolution formula faster, but it has the advantage that
it more accurately reduces to the actual PSF shape as the source's size
decreases.  Decreasing the numerical difference between a small extended
source and the PSF is important when searching for smaller extended
sources and when determining the statistical significance of a detection.

There is one final approximation which speeds up the convolution.
Generally, the PSF is a function of the angle $\theta$ of the incident
event measured relative to the normal to the LAT. To avoid repeatedly
convolving the source shape with the PSF at a series of $\theta$ angles,
an effective PSF is fit to the weighted sum of the PSF times the exposure
at different $\theta$ angles.  Since the source convolution has to be
done at each fit iteration, these optimization were essential for making
this tool fast enough for the following analysis. Typically, the entire
process of fitting the extension and estimating the errors for a source
would run on the order of an hour.

\subsection{Extension Errors}

Errors on fit parameters are typically calculated in one of two
ways. The first method requires explicilty varying one paramter until
the log of the likelihood has fallen by a value 1/2 while simulatenously
optimizing the other fit parmaeters.  \minuit provides the {\em Simplex}
function to calculate errors this way (\cite{minuit_documentation}).
The second method involves estimating the likelihood function as being
a multivariate gaussian and calcluating the curvature of the log of
the likelihood at the peak to estimate when the function will fall by
a value 1/2. \minuit provides the {\em HESSE} algorithm to calcualte
errors this way.  The {\em HESSE} are significantly faster but make the
approximation that the log of the likelihood is perfectly gaussian.

We found that although much faster, the {\em HESSE} algorithm would
somtimes grosely understimate the errors on the localization and extension
of the source.  On the other hand, we found the {\em Simplex} algorithm
was computationally very slow and would often run into convergence
problems and fail to get reasonable errors.

As a compromise, we employed a hybrid approach to calculate
the localization and extension errors in \pointlike by making the
approximation that the covariance between the extension and location of
the source was small. On the other hand, we do not necessarily expect the
likelihood profile to be Gaussian when varying the source's extension.
So we calculate the error on the extension by fixing the position of the
source and then explicitly using {\em Simplex} to vary the extension of
the source until the log of the likelihood has fallen by 1/2.

Then, the extended source's position error is calculated using the
same code employed to calculate position errors of point sources. This
code is used to calculate the localization error of all point sources
kn the catalog\cite{second_cat}. It works by fitting an ellipse to the
likelihood surface near the maximum and using this fit ellipse to estimate
an elliptical localization errors. By reporting two localization errors
as well as a localization error angle, this method explicitly takes care
of covariance between the two position coordinates.

\todo[inline]{Document the extension profile
shown in figure~\ref{extension_profile}.
Make an extension profile plot.
show it in figures and describe it.
}

\subsection{Gtlike Crosscheck}

Because of its computational speed, \pointlike is especially well
optimized for analysis that requires many iterations. This include
source localization, extension fitting, source finding, building test
statistic maps, and model comparison. On the other hand, \pointlike's
efficiency has to entail some loss of precision and so it is generally
expected that \gtlike will produce a more precise result.  Furthermore,
because it is the standard likelihood analysis package, \gtlike has been
more extensively validated and is very well trusted.

For that reason, we take our best fit source model obtained by
\pointlike with the best fit positions and extensions and refit the
spectral parameters of the model using \gtlike.  \gtlike provides a
second calculation of the likelihood which can then be used to obtain
for any source a second measure of \ts and \tsext.

We find very good agreement between the two methods. For the new extended
source candidates, Table \ref{alt_diff_model_results} shows a comparison
of \ts and \tsext computed using \pointlike and \gtlike. For the rest
of this paper, unless otherwise mentioned all \ts, \tsext, flux, and
spectral index values were obtained using \gtlike with the localization
results found by \pointlike.

@@

There is a degeneracy in the signal expected from spatially extended
sources and multiple point sources.  To assess the possibility of source
confusion, we developed a function in pointlike for simulatenously fitting
two point sources so that the likelihood fitting two point sources
could be compared to the likelihood fitting one extended source.

The algorithm works by first dividing the selected point source into
two separate sources. Then, the position of the two sources 
are fit simultanously. To avoid projection effects, the fit is
done in a coordiante system rotated to the celestial equator.
In this rotated coordiante system, space can be well approximated
as being flat and we are interested in fitting the 
rotated coordiante $(x_1',y_1')$ and $(x_2',y_2')$
We decided not to directly fit the positions of the two sources 
but instead the fit sum and difference of the rotated coordiantes
$S_x=(x_2'+x_1')/2$, $S_x=(y_2'+y_1')/2$, $D_x=(x_2'-x_1')/2$, and $D_y=(y_2'-y_1')/2$.
We believe that fitting the average of position and the difference
independently would cause less degeneracy in the fit and help the fitter
find the best position.

For a particular value of $S_x$, $S_y$, $D_x$, and $D_y$, the Poisson
likelihood is computed by converting the sum and difference coordinates
into the rotated coordiantes, unrotating the coordiantes, moving the
two sources in the \pointlike to the new positions, and performing a new
spectral analysis at the modified position.  The function then maximizes
The Poission likelihood $L(S_x,S_y,D_x,D_y)$ as a function of the sum
and difference of the rotated coordaintes using the Minuit to fit the
sum and differenc coordiantes (\cite{minuit_documentation}).

After the minimization converges, we can define the parameter $\tsext$
as twice the increase in log likelihood between fitting the region with
two point sources and fitting the region with one point source
\begin{equation}
  \tsext=2\times(LL_\text{ext}-LL_\text{ps}).
\end{equation}
Generally, \tsinc can be directly compared to \tsext to see wheter two
point sources are prefered over one extended source. Unfortunatly,
performing a likelihood ratio test between two unnested models
is not a calibrated test and so there is no straightforward way
to convert the change in likelihood into a significance~\ref{
http://arxiv.org/abs/astro-ph/0201547 }. Nevertheless, we find this
ratio, at least in the extreme case where $\tsinc \ll \tsext$ or
$\tsinc\gg\tsext$ to be suggestive. For all the spatally extended
candidates found, we computed both \tsext and \tsinc.

Because of \pointlike's computational efficiency, this algorithm
runs in a reasonable amount of time.

@@

When running \pointlike, we have to pick a spatial model for an extended
source. Our spatial model is then convolved with the PSF to create the
probability density function (PDF) that is is fit to the obersved counts.
Possible choices for spatial models include a Two dimensional gaussian
source (defined as a Gaussian in $x$ times a Gaussian in $y$):
\begin{equation}
  \text{PDF}(x,y)=\tfrac{1}{2\pi\sigma^2}\exp\left(-(x^2+y^2)/2\sigma^2\right)
\end{equation}
or a uniform disk:
\begin{equation}
  \text{PDF}(x,y)=\tfrac{1}{\pi\sigma^2}\delta\left(\sqrt{x^2+y^2}<\sigma\right).
\end{equation}

Although these shapes look significantly different from eachother, the
convolution of these spatial models of typtical sizes with the PSF will
produce PDFs that are very similar.

This is demonstrated in figure~\ref{compare_disk_gauss}.  The plot shows
(in black) the PSF that would be observed for a source of spectral
index 2. The dashed red (blue) line shows the probabability density
function for a Gaussian (uniform disk) of width $0.5\deg$.  The solid
red (blue) line shows the probability density function for the same
Gaussian (uniform disk) convolved with the PSF of an index 2 source.
figure~\ref{compare_disk_gauss} shows the analysis using an energy range
betewen 1 \gev and 100 \gev whereas the right plot shows an analysis using an
energy range from 10 \gev to 100 \gev.

This plot shows that significantly different extended source shapes
look very similar after smearing with the PSF. The convolved shape of
the disk and Gaussian admittedly do look more different when restrict
ourselves to higher energy where the PSF is narrower, but we do not
excpet to have much luch learning about the exact nature of the
extended source shape. For that reason, in this paper we always use
the uniform disk spatial model to emphaizie our ignorance to the true
shape of the extended sources we are studying.

@@

Finally, we note that comparing observed morphologies to 
counts maps can be misleading.  This effect can best be seen
in figure~\ref{compare_r68} where we have simulted a
extended source with 
a spatial model that is a uniform disk of extension 0.5\deg,
integral flux between 1 \gev and 100 \gev of $3e-8\ph/\cm^2/s$,
and a spectral index of 2.

To reduce Poisson noise, figure~\ref{compare_r68} shows a counts map of
this simulated source smoothed with a 0.1\deg Gaussain kernel.  Overlaied
on the plot is the results of fitting this simulated source with a uniform
disk and a Gaussian spatial model.  The solid white line represents the
fit extension $\sigma_\text{disk}=0.496$ for the disk which was found to
be in very good agreemen with the simulated value.  The solid black line
represenst the fit extension $\sigma$ for the Gaussian which was found to
be $\sigma_\text{Gaussian}=0.2627$.  The white dashed line is the 68\%
containment radius of the fit Disk shape and the black dashed line is
the 68\% containment radius of the fit Gaussian shape.  Even though the
fit $\sigma$ for the disk and uniform Gaussian are different, the fit
68\% continment radius are very similar ($\rsixeight_\text{,disk}=0.41$
whereas $\rsixeight_\text{,Gaussian}=0.40$).

@@

The second conclusion we make is that because a counts map is smeared
out by the PSF and depends strongly on the particular color bar used, it
is very difficult to determine directly from a counts map a reasonable
spatial model that fits the data.  It is much better to obtain the
spatial model by a direct fit of a spatial parameters of a selected
model to the data.  The third conclusion we make is that when comparing
fit extensions to those at other wavelenghts (espeically in \tev), it is
important to compare fits of consistent spatial models. If this is not
possible, it is impotant to at least compare indirectly the $rsixeight$
for the different spatial models.

@@

\subsection{Significance of Extension}

>>>>

% Note, hyperref has to come after other packages!
\usepackage{hyperref}

@@

\begin{abstract}
We present a new method for analyzing spatially extended sources with
the Fermi Large Area Telescope (LAT), the primary science instrument
on the {\em Fermi Gamma-ray Space Telescope (Fermi)}.  We provide a
series of Monte Carlo studies that validate this tool.
We then calculate the Fermi LAT's sensitivity to spatially extended
sources.  We then use this tool to test all sources in the two
year source catalog for extension and report on the discovery
of XXXXXX\todo{What number should be put
here???} new extended sources.
\end{abstract}

@@

The Large Area Telescope (LAT) is a pair conversion telescope on the
Fermi Gamma-Ray Space Telescope. It has been surveying the $\gamma$-ray
sky since June 2008.  Fermi is well suited as a survey
instrument because of its broad energy coverage (20 \mev
to $>300$ \gev), wide field of view ($\sim 2.4 \sr$) and large effective
area ($\sim 8000 \cm^2$ at $>1 \gev$).

Using one year of all sky surveying data, the LAT published a catalog of
source that were significantly detected.  Many of these were associated
with a variety of source classes including active galactic nuclei,
pulsars, galaxies, and supernova remnants.  Many of these classes can be
spatially resolved when observed at other frequencies and could possibly
be spatially resolvable in \gev by the LAT.

The LAT collaboration has previously reported on the discovered
a variety of spatially extended supernova remnants (SNRs)
IC443 and W51C, pulsar wind nebulae (PWNe) MSH 15-52 and Vela X
(\cite{ic443,w51c,msh1552,velax}), and galaxies the Large Magellanic
Cloud and the Small Magalenic Cloud (\cite{lmc,smc}.  There are other
source classes which could possibly be spatially resolved by the LAT
including galaxy clusters, dark matter, and possibly new physics.

Furthermore, a variety of spatially extended but otherwise unassociated
sources have been detected in the galactic plane at \tev energies using
Air Cherenkov detectors. Most prominent was a survey of the galactic plane
using the High Energy Stereoscopic System (H.E.S.S) which discovered 14
spatially extended sources (\cite{HESS_plane_survey}).

Being able to spatially resolve a LAT source is important for several
reasons. Most importantly, the spatial morphology can be used to
help identify the source at other wavelengths. Because of the large
point spread function (PSF), there are often several possible source
counterparts.  Finding a matching morphology can uniquely identify
the source.

Because of the largely varying PSF of the LAT, the spatial and spectral
model of a source do not nicely decouple. Therefore, spatially resolving
a LAT source is also important for obtaining an unbiased spectrum.
Furthermore, correctly modeling a source's extension is important for
improved the overall model of the sky and removing excess residuals.
These residuals around catalog sources can lead to finding spurious nearby
point sources. This was found to be the case with the one year catalog
in the region surrounding the Large Magellanic Cloud (\cite{first_cat}).

@@

Morphological studies of sources in the \gev energy range
using the Fermi LAT are challenging because of the significantly energy
dependent PSF and because of systematic errors in the galactic diffuse
emission.  The LAT's PSF is determined at lower energies by multiple
scattering in the tracker and the 68\% containment radius of the PSF is
approximately 5.1\deg at 100 \mev  when averaged over instrument acceptance
and including photons which convert in either the thick or thin layers of
the tracker. The PSF approaches a limit given by the granularity of the
tracker channels and is 0.14\deg at 10 \gev (\cite{on_orbit_calibration}).
Therefore, the higher energy photons are significantly more important for
resolving a source but are limited by statistics. Therefore,
a detailed analysis procedure mused be used to maximize our
sensitivity.

@@

A new analysis tool has been developed to address the unique requirements
for studying spatially extended sources with the Fermi LAT. The tool
provides a maximum likelihood analysis where the Poisson likelihood
for observing the measured counts is computed is maximized by fitting
parameters of the model.  The extension of a source can be modeled by a
geometric shape (e.g. a disk or Gaussian) and the source's position
and extension can simultaneously fit by maximizing the likelihood.

This analysis is not very feasible using the official LAT likelihood
analysis tool \gtlike\footnote{\gtlike is distributed publicly by the
Fermi Science Support Center\cite{fssc}} because \gtlike is only able to
fit the spectral parameters of the source. The extension fitting would
have to be done as an external loop outside of the code. Typically, the
run time for a binned source analysis using \gtlike is several hours
so this loop would be especially time consuming. What has typically
been done to study an extended sources is to fix the position of the
extended source and then develop a profile of the likelihood as a
function of the extension.  This approach is not optimal because the
position and extension of the source should be simultaneously maximized
to find the best fit parameters.  Furthermore since this approach is
very computationally intensive, no large scale Monte Carlo effort can
be launched to validate it.

An alternate approach was developed for fitting spatially extended
sources.  An alternate maximum likelihood fitting package called
\pointlike has been developed by the Fermi collaboration and we have
added to it the ability to analyze spatially extended sources. \pointlike
is described in the one year catalog and in Matthew Kerr's Ph.D thesis
(\cite{first_cat,matthew_kerr_thesis}).

What differentiates \pointlike most significantly from \gtlike is its
emphasis on speed by introducing approximations into the calculation
of the likelihood function.  This is done while at the same time
trying to minimized the introduced numerical errors.  Like the binned
version of \gtlike, \pointlike bins the sky both in position and energy.
Unlike \gtlike, \pointlike relies on the {\em healpix} representation of
spatial bins (\cite{healpix}). Since the PSF changes significantly with
energy, the spatial size of each bin for a given energy is calculated
to be smaller than than the PSF. This ensures that little information
is lost at each energy level while minimizing the overall number of bins.

\pointlike also bins $\gamma$-rays depending upon whether they
convert in the front or back parts of the detector. Since the PSF
is smaller for front entering photons, binning each conversion type
independently improves the tool's sensitivity.
Furthermore, \pointlike uses a sparse matrix representation of the spatial
bins where only the bins with counts in them are used. \pointlike
then independently integrates the overall model predicted counts in the region.

The intelligent binning and sparse matrix representation of the data
effectively interpolates between a binned analysis at low energy (where
the resolution is poor and there are many counts) and an unbinned
analysis at high energy (where the resolution is good but there are
few counts). This approach provides a significant savings in time while
introducing small numerical error.

Finally, \pointlike is written mostly in the \python programming
language and uses {\em numpy} and {\em scipy} to vectorize most of the
calculations. This provides a reasonable trade off between computational
efficiency and ease of development. By leaving the interface as a series
of \python modules, advanced functions including extension fitting
(described in section~\ref{extesion_fitting}) and a dual localization
(described in section~\ref{dual_localization_method}) could be quickly
added on top of the core functionality.

@@

We modified \pointlike to allow modeling the spatial extension of
of a source. It is assumed that the spatial and spectral information
completely decouple and that the spatial shape can be represented by a
simple geometric function.  The code consistently convolves the extended
source shape with the PSF (as a function of energy) when computing the
model predicted counts in a bin.

\pointlike uses the \minuit fitting library to simultaneously maximize
the likelihood by fitting the position and extension of an extended
source (\cite{minuit_documentation}).  For each position and extension,
the spectral parameters of the sky model are refit. To avoid projection
effects, what is directly fit is not the longitude and latitude of
the source but instead the displacement in a rotated reference frame.
The significance of the extension of a source can be calculated from
\tsext, which is defined as twice the increase in log likelihood between
the extended source model at its best fit values and the
point source model at its best fit values:
\begin{equation}
  \tsext=2\times(LL_\text{ext}-LL_\text{ps}).
\end{equation}
The interpretation of \tsext in terms of a statistical significance is
discussed in section~\ref{significance_of_extension}.

The extended source code is speed up in two ways. First, the convolution
of the spatial shape of a radially symmetric extended source with the PSF is
done semi-analytically. The observed photon
distribution can be written as the convolution of the source shape ($I_\text{src}$)
with the psf
\begin{equation}
  \text{PDF}(\vec r) = \int  \text{PSF}(|\vec r - \vec r'|)I_\text{src}(\vec r') d A'.
\end{equation}
The PSF is parameterized by
\begin{equation}
  \text{PSF}(\vec r) = 
  \frac{1}{2\pi\sigma^2}
  \left(1-\frac{1}{\gamma}\right)
  \left(1+\frac{u}{\gamma}\right)^{-\gamma},
\end{equation}
where $\sigma$ and $\gamma$ have been found empirically as a function
of energy.  For radially symmetric
extended sources, the angular part of the integral can be done analytically
leaving a easier one dimensional integral:
\begin{equation}
  \text{PDF}(u)= \int_0^\infty dv
  I_\text{src}(v) 
  \left(\frac{\gamma-1}{\gamma}\right)
  \left( \frac{\gamma}{\gamma + u + v}\right)^\gamma 
  \times ~_2F_1 \left(\gamma/2,\frac{1+\gamma}{2},1,\frac{4uv}{(\gamma+u+v)^2}\right).
\end{equation}
This convolution formula is faster and more closely reduces to the PSF
when the source's size is very small. Decreasing this the numerical
error is important when evaluating the statistical significance of a
source's extension.

The other approximation is to ignore the dependence of the PSF on the
incident angle of a $\gamma$-ray.  Generally, the PSF is a function
of the incident angle of a photon measured with respect to the LAT.
To avoid multiple convolutions of the source shape with the PSF, an
effective PSF is calculated by weighted the PSF for different incident
angles.  Since the convolution has to be done at each iteration of the
extension fit, these approximations were necessary to make the tool run
in a reasonable amount of time.

\subsection{Extension Errors}
\label{extension_error}

Errors on fit parameters are typically calculated in one of two
ways. The first method requires explicitly varying one parameter until
the log of the likelihood has fallen by a particular value while 
simultaneously
optimizing the other fit parameters. The
\minuit fitting package provides the {\em Simplex}
function to calculate errors this way (\cite{minuit_documentation}).
The second method involves estimating the likelihood function as being
a multivariate Gaussian and calculating the curvature of the log of
the likelihood at the peak to estimate when the function will fall 
by the same value. \minuit provides the {\em HESSE} algorithm to calculate
errors this way.  The {\em HESSE} errors are significantly faster but 
are only approximately true.

We found that the {\em HESSE} algorithm would sometimes badly
underestimate the errors on the extension of the source.  On the other
hand, the {\em Simplex} algorithm was prohibitively slow and would often
fail to convergence. \pointlike's approach to calculating extension errors
is to fixing all the spatial parameters and varying just the extension
until the likelihood has fallen by the desired amount.  This approach
makes the approximation that the covariance between each spatial parameter
is small but not that the likelihood is Gaussian. This method 
is shown schematically in figure~\ref{extension_profile} which shows
the change in log likelihood when varying the fit extension of
the SNR IC443.

The localization error is then calculated by fixing the extension to the
best fit value and fitting an ellipse to the likelihood as a function of
the spatial parameters.  This fit can be used to calculate an elliptical
localization error which correctly accounts for covariance between the
two position coordinates.  This same method is used to calculate the
localization error of all point sources in the LAT catalog
(\cite{first_cat}).

\subsection{Gtlike Crosscheck}
\label{gtlike_crosscheck}

Because of its computational speed, \pointlike is important for analysis
that require many iterations such as source localization and extension
fitting.  On the other hand, we expect \gtlike to get a slightly
more precise answer for fitting the spectral parameters of a source.
Furthermore, because \gtlike is the standard likelihood analysis package,
it has been more extensively validated for spectral analysis.

For that reason, in the following analysis we took the best fit
positions and extensions of the source model fit by \pointlike but
refit the spectral spectral parameters of the model using \gtlike.
\gtlike provides a second calculation of the likelihood which gives a
second measure of \ts and \tsext.

We found very good agreement between the two methods. For the new
extended source candidates, table~\ref{alt_diff_model_results} shows a
comparison of \ts and \tsext computed using \pointlike and \gtlike. For
the rest of this paper, unless otherwise stated all \ts, \tsext, and
fit spectral parameters were found using \gtlike with the localization
results computed by \pointlike.

@@

There is a degeneracy between a spatially extended source and multiple
point sources.  To assess the possibility of source confusion,
we developed a function in \pointlike to simultaneously fit two point
sources and compare the likelihood of two point sources to the likelihood
of one extended source.

The code uses \minuit to simultaneously fit the position of the two point
sources.  To avoid projection effects, the fit is done in a coordinate
system rotated to the celestial equator.  Instead of directly fitting
the rotated coordinates $x_1$ and $x_2$, we decided there would be less
covariance by fitting instead the sum and difference of the rotated
coordinates $S=(x_1+x_2)/2$ and $D=(x_2-x_1)/2$.  The log likelihood
$L(S,D)$ is maximized by fitting the sum and difference of the two point
sources.  Because of \pointlike's computational efficiency, this algorithm
runs in a reasonable amount of time. Similar to extension described in
section~\ref{gtlike_crosscheck}, the spectrum of can be refit using \gtlike
at the best fit positions.

We can define the parameter \tsinc as twice the increase in log likelihood
fitting the region with two point sources compared to fitting the region
with one point source:
\begin{equation}
  \tsinc=2\times(LL_\text{2pts}-LL_\text{ps}).
\end{equation}
A straightforward likelihood ratio test between the two point source
model and the extended source model because the models are note nested.

\tsinc can not be directly compared to \tsext to see whether two
point sources are preferred over one extended source. Unfortunately,
performing a likelihood ratio test between two models that are not nested
(\cite{statistics_with_care}). So the comparison of \tsext with \tsinc is
not a calibrated test.  Nevertheless, we find the extreme cases where
$\tsinc \ll \tsext$ or $\tsinc\gg\tsext$ to be suggestive and use this
as a test for source confusion. For the spatially extended candidates
found, we computed both \tsext and \tsinc.  The results are presented
in table~\ref{dual_localization_results}

@@

An extended source must be modeled by an assumed spatial shape and
the convolution of that shape with the PSF is fit to the data.
Possible choices for spatial models include a two dimensional Gaussian
\begin{equation}
  \text{PDF}(x,y)=\tfrac{1}{2\pi\sigma^2}\exp\left(-(x^2+y^2)/2\sigma^2\right)
\end{equation}
or a uniform disk:
\begin{equation}
  \text{PDF}(x,y)=\tfrac{1}{\pi\sigma^2}\delta\left(\sqrt{x^2+y^2}<\sigma\right).
\end{equation}
Although these shapes are significantly different from each other, the
convolution of these spatial models of typical sizes with the PSF will
produce PDFs that are very similar.

This is demonstrated in figure~\ref{compare_disk_gauss}.  The plot shows
(in black) the PSF that would be observed for a source of spectral
index 2. The dashed red (blue) line shows the probability density
function for a Gaussian (uniform disk) of width $0.5\deg$.  The solid
red (blue) line shows the probability density function for the same
Gaussian (uniform disk) convolved with the PSF of an index 2 source.
Figure~\ref{compare_disk_gauss} shows the analysis using an energy range
between 1 \gev and 100 \gev whereas the right plot shows an analysis using an
energy range from 10 \gev to 100 \gev.

This plot shows that significantly different extended source shapes
look very similar after smearing with the PSF. The convolved shape of
the disk and Gaussian admittedly do look more different when restrict
ourselves to higher energy where the PSF is narrower, but we do not
except to have much luck learning about the exact nature of the
extended source shape. For that reason, in this paper we always use
the uniform disk spatial model to emphasize our ignorance to the true
shape of the extended sources we are studying.

@@

Finally, we note that comparing observed morphologies to 
counts maps can be misleading.  This effect can best be seen
in figure~\ref{compare_r68} where we have simulated a
extended source with 
a spatial model that is a uniform disk of extension 0.5\deg,
integral flux between 1 \gev and 100 \gev of $3e-8\ph/\cm^2/s$,
and a spectral index of 2.

To reduce Poisson noise, figure~\ref{compare_r68} shows a counts map of
this simulated source smoothed with a 0.1\deg Gaussian kernel. Overlaid
on the plot is the results of fitting this simulated source with a uniform
disk and a Gaussian spatial model.  The solid white line represents the
fit extension $\sigma_\text{disk}=0.496$ for the disk which was found to
be in very good agreement with the simulated value.  The solid black line
represents the fit extension $\sigma$ for the Gaussian which was found to
be $\sigma_\text{Gaussian}=0.2627$.  The white dashed line is the 68\%
containment radius of the fit Disk shape and the black dashed line is
the 68\% containment radius of the fit Gaussian shape.  Even though the
fit $\sigma$ for the disk and uniform Gaussian are different, the fit
68\% containment radius are very similar ($\rsixeight_\text{,disk}=0.41$
whereas $\rsixeight_\text{,Gaussian}=0.40$).

@@

The second conclusion we make is that because a counts map is smeared
out by the PSF and depends strongly on the particular color bar used, it
is very difficult to determine directly from a counts map a reasonable
spatial model that fits the data.  It is much better to obtain the
spatial model by a direct fit of a spatial parameters of a selected
model to the data.  The third conclusion we make is that when comparing
fit extensions to those at other wavelenghts (especially in \tev), it is
important to compare fits of consistent spatial models. If this is not
possible, it is impotant to at least compare indirectly the $rsixeight$
for the different spatial models.

@@

\subsection{Significance of Extension}
\label{significance_of_extension}

The value of \tsext, the increase in likelihood fitting the source
with an extended spatial model, can be used to calculate
the statistical significance of the source's extension.
Since the point and extended models are nested hypothesis,
we would always expect $\tsext>0$.

We would expect that if we were really testing a point source that
\tsext would not be very large since a point model would fit about
as good as anytthing else. Therefore, we can consider a source to be
extended (and reject the null hyopthesis) if \tsext is large enough.
Since going from fitting a point soue to fitting a simple extended
source introduces only one degree of freedom (the source's extension)
Wilks' Theorem says that the distribution of $\tsext$ wen testin
point sources should follow a $\chi^2_1$ distribution with one degree
of freedom (\cite{wilks_theorem}).
